# Overview

This document contains detailed descriptions of the suplementary code for [@manuscript]. Three case studies are presented along with a simulation study.

```{r}
#| label: quarto-setup
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 12,
  fig.height = 9,
  fig.align = "center",
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
options(progressr.enable = FALSE)
```

## Case Studies

This repository demonstrates MrIML 2.0 across four ecological systems. Each case study is self-contained with data, code, and interpretation.

### Case Study 1: Avian Blood Parasites

Analysis of co-occurrence patterns in vector-borne parasites from New Caledonian birds. Demonstrates asymmetric associations and the relative importance of host abundance versus biotic interactions.

**Key findings:** Host abundance is the primary driver, but parasite co-occurrence significantly improves predictions. Asymmetric associations reveal directional effects between parasite species.

### Case Study 2: Forest Communities

Multi-taxonomic analysis of beetles, birds, and plants across forest successional gradients. Compares assembly mechanisms among different taxonomic groups.

**Key findings:** Environmental filtering dominates across all groups. Plant communities show strongest responses to successional gradients, while bird and beetle communities respond primarily to altitude and microclimate.

### Case Study 3: Ostrich Gut Microbiome

Microbiome development across three gut sections (ileum, caecum, colon) in juvenile ostriches. Examines age-dependent assembly and section-specific patterns.

**Key findings:** Critical developmental window at 20-40 days post-hatching. Positive associations dominate across all gut sections, with age and weight as primary predictors.

## Simulation study

A simulation study.

## Computing enviroment information

```{r}
#| echo: false
#| output: false
# Load required packages
# Data manipulation
library(tidyverse)
library(DT)
library(readxl)
library(here)
# Modeling
library(mrIML)
library(tidymodels)
library(future)
library(finetune)
library(flashlight)
# Plotting
library(igraph)
library(ggnetwork)
library(cowplot)
library(patchwork)
```

```{r}
#| label: session-info
sessionInfo()
```
