# Avian blood parasites

```{r}
#| label: setup-case-1
#| message: false
#| warning: false
#| cache: false
# Load required packages
# Data manipulation
library(tidyverse)
library(DT)
library(here)
# Modeling
library(mrIML)
library(tidymodels)
library(future)
library(finetune)
library(flashlight)
# Plotting
library(igraph)
library(ggnetwork)
library(cowplot)
library(patchwork)

# Load custom functions
source(here("R-functions", "select_variables.r"))
source(here("R-functions", "plot_network.r"))

# Set up parallel processing
n_cores <- parallel::detectCores()
plan("multicore", workers = 2)

# Set seed
set.seed(1234)
```

This case study demonstrates the application of *MrIML 2.0* to a small dataset of vector-borne blood parasites in New Caledonian birds. We analyze presence/absence data from four parasite species:

-   *Haemoproteus zosteropis* species complex
-   *H. killangoi* species complex\
-   *Plasmodium* spp.
-   Nematode *Microfilaria* species

The dataset includes 449 individual birds and a covariate representing the proportion of captures attributed to *Zosterops* spp. at each site, serving as a proxy for host relative abundance.

```{r}
#| label: load-bird-parasite-data
data <- MRFcov::Bird.parasites
datatable(data)
```

To prepare the data for multi-response modeling, we separate it into `Y` (response), `X` (predictors), and `X1`(co-occurrence predictors). This is how data is parsed to `mrIML`.

```{r}
#| label: data-prep-bird-parasite-data
# Prepare response variables (parasite species)
Y <- data %>%
  dplyr::select(-scale.prop.zos) %>%
  dplyr::select(sort(names(.)))

# Prepare predictor variables (host abundance proxy)
X <- data %>%
  dplyr::select(scale.prop.zos)

# Co-occurrence data (other parasites as predictors)
X1 <- Y %>%
  dplyr::select(sort(names(.)))
```

## Model definition

We formulate three models of community assembly to compare the relative importance of co-occurrence versus environmental effects:

1.  **Parasite co-occurrence only** (just X₁, no other covariates)
2.  **Host abundance only** (just X, the 'host abundance model')
3.  **Combined model** (both X₁ and X, co-occurrence + host abundance)

However, all three of these model configerations utilise the same base random forest model specified using `tidymodels`.

```{r}
#| label: model-def
# Define random forest model
model_rf <- rand_forest(
  trees = 1000,
  mode = "classification",
  mtry = tune(),
  min_n = tune()
) %>%
  set_engine("randomForest")
```

### Model 1: Combined model (co-occurrence + host abundance)

```{r}
#| label: model-combined
#| results: "hide"
yhats_rf_combined <- mrIMLpredicts(
  X = X,
  Y = Y,
  X1 = X1,
  Model = model_rf,
  balance_data = 'no',
  tune_grid_size = 5,
  prop = 0.7,
  k = 5,
  racing = TRUE
)
```

### Model 2: Host abundance only

```{r}
#| label: model-host-only
#| results: 'hide'
yhats_rf_nox1 <- mrIMLpredicts(
  X = X,
  Y = Y,
  X1 = NULL,
  Model = model_rf,
  balance_data = 'down',
  tune_grid_size = 5,
  prop = 0.7,
  k = 5,
  racing = TRUE
)
```

### Model 3: Co-occurrence only

```{r}
#| label: model-cooccur-only
#| results: 'hide'
yhats_rf_nox <- mrIMLpredicts(
  X = NULL,
  Y = Y,
  X1 = X1,
  Model = model_rf,
  balance_data = 'down',
  tune_grid_size = 5,
  prop = 0.7,
  k = 5,
  racing = TRUE
)
```

## Model performance

The proformance of each model for the diferen taxa can be explored using `mrIMLperformance()`. Below is an example for the combined model.

```{r}
ModelPerf_combined <- mrIMLperformance(yhats_rf_combined)
ModelPerf_combined$model_performance
```

We compare the three model configerations against one another based on average AUC.

```{r}
# Evaluate model performance
ModelPerf_noX1 <- mrIMLperformance(yhats_rf_nox1)
ModelPerf_noX <- mrIMLperformance(yhats_rf_nox)

tibble(
  Model = c(
    "Combined model",
    "Host abundance only",
    "Co-occurrence only"
  ),
  AUC = c(
    ModelPerf_combined$model_performance$roc_AUC %>%
      mean(),
    ModelPerf_noX1$model_performance$roc_AUC %>%
      mean(),
    ModelPerf_noX$model_performance$roc_AUC %>%
      mean()
  ) %>%
    round(., 3)
)
```

The combined model shows improved performance over the individual components, suggesting that both co-occurrence patterns and host abundance contribute to parasite distribution patterns, however this is a small dataset and so there is significant variation in the metric depending on the cross-validation and test-training split.

## Bootstrap Analysis for Uncertainty Estimation

From here on, we focus on the combined model. However, we can use bootstraping to capture the uncertainty.

```{r}
#| results: "hide"
# Run bootstrap analysis for uncertainty estimation
bs_avian <- yhats_rf_combined %>%
  mrBootstrap(num_bootstrap = 100)
```

```{r}
# Calculate variable importance with bootstrap confidence intervals
bs_impVI <- mrVip(
  yhats_rf_combined,
  mrBootstrap_obj = bs_avian,
  threshold = 0.0,
  global_top_var = 10,
  local_top_var = 5,
  model_perf = ModelPerf_combined
)

# Display variable importance plot
bs_impVI[[3]]
```

## Partial Dependence Analysis

```{r partial-dependence}
# Create bootstrapped partial dependence plots for Plasmodium
pds <- mrPdPlotBootstrap(
  yhats_rf_combined,
  mrBootstrap_obj = bs_avian,
  vi_obj = bs_impVI,
  target = 'Plas',
  global_top_var = 5
)

# Display partial dependence plots
pds[[2]]
```

## Environmental Covariate Effects

```{r covariate-effects}
# Examine the relationship between host abundance and parasite occurrence
covar <- mrCovar(
  yhats_rf_combined,
  var = 'scale.prop.zos',
  sdthresh = 0.01
)

# Display the three components of covariate analysis
covar[[1]] / covar[[2]] / covar[[3]]
```

## Interaction Analysis

```{r interaction-analysis}
# Analyze feature interactions using H-statistics
int_ <- mrInteractions(
  yhats_rf_combined,
  num_bootstrap = 100,
  feature = 'Plas',
  top_int = 10
)

# Display interaction plots
int_[[1]] / int_[[2]] / int_[[3]]
```

## SHAP Analysis for Individual Predictions

```{r shap-analysis}
#| results: "hide"
# Perform SHAP analysis for detailed feature attribution
shap_results <- mrShapely(
  yhats_rf_combined,
  taxa = "Plas",
  plot_feature_effects = TRUE,
  plot_dependencies = TRUE,
  plot_2D_dependencies = FALSE
)
```

```{r}
# Display SHAP results
p <- (shap_results$feature_effects$Plas[[1]] + theme_minimal()) +
  (shap_results$feature_effects$Plas[[1]] + theme_minimal()) +
  plot_annotation("Plas") +
  plot_layout(guides = "collect", axis_titles = "collect")
```

## Co-occurrence Network Construction

```{r}
# Build co-occurrence network from bootstrap results
assoc_net <- mrCoOccurNet(bs_avian)

plot_network(
  assoc_net,
  strength_threash = 0.05,
  network_title = "Parasite Co-occurrence Network"
)
```