# Co-occurance simulations

```{r}
#| message: false
#| warning: false
#| results: "hide"
# Load required packages
# Data manipulation
library(here)
library(readxl)
library(tidyverse)
library(DT)
# Modeling
library(mrIML)
library(tidymodels)
library(future)
library(finetune)
library(flashlight)
# Plotting
library(igraph)
library(ggnetwork)
library(cowplot)
library(patchwork)

library(seqtime)
library(Hmsc)

# Load custom functions
source(here("R-functions", "select_variables.r"))
source(here("R-functions", "plot_network.r"))

# Set random seed
set.seed(109)

# Set up parallel processing
#n_cores <- parallel::detectCores()
#plan("multisession", workers = n_cores - 2)
plan("sequential")

# Load simulated data
source(
  here("sim-study", "symetric-sims.r")
)
```

Generate a co-ocurance matrix, M.

```{r}
network_size <- 20
k_average <- 4

simulated_network <- generateM_specific_type(
  nn = network_size,
  k_ave = k_average,
  type.network = "random",
  type.interact = "random",
  interact.str.max = 0.4,
  mix.compt.ratio = 0.5
)
true_network <- simulated_network[[1]]
M <- simulated_network[[2]]
```

plot population dynamics

```{r}
# initial abundance
y <- rpois(network_size, lambda = 100)
# growth rates
r <- runif(network_size)
res <- glv(network_size, M, r, y)
tsplot(10 * res[, 20:1000], time.given = T)
```

Generate dataset on species abundance using the GLV model and convert to presence-absence data.

```{r}
data <- generateDataSet(
  900,
  M,
  count = network_size * 10000,
  mode = 4
) %>%
  t() %>%
  as.data.frame() %>%
  mutate(
    across(everything(), ~ ifelse(. < 1, yes = 0, no = 1))
  )
```

Prepare data and fit mrIML model.

```{r}
#| results: 'hide'
Y <- filterRareCommon(data, lower = 0.01, higher = 0.99)
X1 <- Y

model_rf <- rand_forest(
  trees = 100,
  mode = "classification",
  mtry = tune(),
  min_n = tune()
) %>%
  set_engine("randomForest")

yhats_rf_sim <- mrIMLpredicts(
  Y = Y,
  X = NULL,
  X1 = X1,
  Model = model_rf,
  prop = 0.7,
  k = 5,
  racing = FALSE
)
```

Perform bootstrapping.

```{r}
#| results: 'hide'
bs_sim <- mrBootstrap(yhats_rf_sim)
```

Extract co-occurance network.

```{r}
assoc_net <- mrCoOccurNet(bs_sim)
```

Plot the network alongside the truth.

```{r}
mrIML_mat <- assoc_net %>%
  filter(mean_strength > 0.1) %>%
  mutate(
    mean_strength_dir = ifelse(
      direction == "negative",
      yes = -mean_strength,
      no = mean_strength
    ),
    across(contains("taxa"), ~ sub("^sp", "", .))
  ) %>%
  graph_from_data_frame(
    directed = FALSE,
    vertices = sub("^sp", "", names(Y))
  ) %>%
  get.adjacency(attr = "mean_strength_dir", sparse = FALSE)

#plots
g_mrIML <- graph_from_adjacency_matrix(
  mrIML_mat,
  mode     = "undirected",
  weighted = T,
  diag     = F
)

edge_colors_mrIML <- ifelse(E(g_mrIML)$weight < 0, yes = "red", no = "blue")
edge_widths_mrIML <- abs(E(g_mrIML)$weight) * 10 # Adjust the scaling factor as needed
matching_indices <- as.numeric(rownames(mrIML_mat))

true_mat <- M[matching_indices, matching_indices]
rownames(true_mat) <- rownames(mrIML_mat)
colnames(true_mat) <- rownames(mrIML_mat)
g_true <- graph_from_adjacency_matrix(
  true_mat,
  mode = 'undirected',
  weighted = T,
  diag = F
)
edge_colors_true <- ifelse(E(g_true)$weight < 0, yes = "red", no = "blue")
edge_widths_true <- abs(E(g_true)$weight) * 5 # Adjust the scaling factor as needed

layout1 <- layout_nicely(g_true) # This produces a warning but not a big deal
```

Fit HMSC model and extract residual species associations.

```{r}
#| results: 'hide'
studyDesign <- data.frame(sample = factor(1:nrow(Y)))
rL <- HmscRandomLevel(units = studyDesign$sample)
m_hmsc <- Hmsc(
  Y           = as.matrix(Y),
  XFormula    = ~1,
  XData       = data.frame(intercept = rep(1, nrow(Y))),
  distr       = "probit",
  studyDesign = studyDesign,
  ranLevels   = list("sample" = rL)
)
m_hmsc <- sampleMcmc(
  m_hmsc,
  thin      = 50,
  samples   = 1000,
  transient = 5000,
  nChains   = 4,
  verbose   = 0
)
```

```{r}
mpost <- convertToCodaObject(m_hmsc)
psrf <- gelman.diag(mpost$Omega[[1]], multivariate = FALSE)$psrf
psrf
```

```{r}
#| results: 'asis'
if (any(psrf[, "Point est."] > 1.1)) {
  cat("::: {.callout-warning}\n")
  cat("**HMSC convergence concern:** One or more Omega parameters have Rhat > 1.1,")
  cat(" indicating incomplete chain mixing. Interpret these results cautiously.\n")
  cat(":::\n")
}
```

```{r}
assoc <- computeAssociations(m_hmsc)[[1]]
OmegaCor <- assoc$mean
OmegaCor[assoc$support > 0.05 & assoc$support < 0.95] <- 0
hmsc_species_names <- paste0("sp", rownames(mrIML_mat))
hmsc_mat <- OmegaCor[hmsc_species_names, hmsc_species_names]
rownames(hmsc_mat) <- rownames(mrIML_mat)
colnames(hmsc_mat) <- rownames(mrIML_mat)

g_hmsc <- graph_from_adjacency_matrix(
  hmsc_mat,
  mode     = "undirected",
  weighted = TRUE,
  diag     = FALSE
)
edge_colors_hmsc <- ifelse(E(g_hmsc)$weight < 0, yes = "red", no = "blue")
edge_widths_hmsc <- abs(E(g_hmsc)$weight) * 10

par_old <- par(mfrow = c(1, 3))

plot(
  g_mrIML,
  layout = layout1,
  edge.color = edge_colors_mrIML,
  edge.width = edge_widths_mrIML,
  isolates = TRUE,
  main = "mrIML co-occurrence"
)

plot(
  g_hmsc,
  layout = layout1,
  edge.color = edge_colors_hmsc,
  edge.width = edge_widths_hmsc,
  isolates = TRUE,
  main = "HMSC co-occurrence"
)

plot(
  g_true,
  layout = layout1,
  edge.color = edge_colors_true,
  isolates = TRUE,
  edge.width = edge_widths_true,
  main = "Simulated (truth)"
)

par(par_old)
```

Compare the learned networks with the truth.

```{r}
dist_mat_true <- 1 - true_mat
dist_mat_mrIML <- 1 - mrIML_mat
mantel_mrIML <- vegan::mantel(
  as.dist(dist_mat_true),
  as.dist(dist_mat_mrIML),
  method = "spearman"
)
mantel_mrIML

dist_mat_hmsc <- 1 - hmsc_mat
mantel_hmsc <- vegan::mantel(
  as.dist(dist_mat_true),
  as.dist(dist_mat_hmsc),
  method = "spearman"
)
mantel_hmsc
```

## Mutualistic

```{r}
simulated_mutual_network <- generateM_specific_type(
  nn = network_size,
  k_ave = k_average,
  type.network = "random",
  type.interact = "mutual",
  interact.str.max = 0.3,
  mix.compt.ratio = 0.5
)
true_mutual_network <- simulated_mutual_network[[1]]
M_mutual <- simulated_mutual_network[[2]]
```

```{r}
data_mutual <- generateDataSet(
  900,
  M_mutual,
  count = network_size * 10000,
  mode = 4
) %>%
  t() %>%
  as.data.frame() %>%
  mutate(
    across(everything(), ~ ifelse(. < 1, yes = 0, no = 1))
  )
```

```{r}
#| results: 'hide'
# Prepare data
Y_mutual <- filterRareCommon(data_mutual, lower = 0.01, higher = 0.99)
X1_mutual <- Y_mutual
# Fit mrIML obj
yhats_rf_sim_mutual <- mrIMLpredicts(
  Y = Y_mutual,
  X = NULL,
  X1 = X1_mutual,
  Model = model_rf,
  prop = 0.7,
  k = 5,
  racing = FALSE
)
# Bootstrapping
bs_sim_mutual <- mrBootstrap(yhats_rf_sim_mutual)
# Extract network
assoc_net_mutual <- mrCoOccurNet(bs_sim_mutual)
```

```{r}
mrIML_mat_mutual <- assoc_net_mutual %>%
  filter(mean_strength > 0.1) %>%
  mutate(
    mean_strength_dir = ifelse(
      direction == "negative",
      yes = -mean_strength,
      no = mean_strength
    ),
    across(contains("taxa"), ~ sub("^sp", "", .))
  ) %>%
  graph_from_data_frame(
    directed = FALSE,
    vertices = sub("^sp", "", names(Y_mutual))
  ) %>%
  get.adjacency(attr = "mean_strength_dir", sparse = FALSE)

#plots
g_mrIML_mutual <- graph_from_adjacency_matrix(
  mrIML_mat_mutual,
  mode = 'undirected',
  weighted = T,
  diag = F
)

edge_colors_mrIML_mutual <- ifelse(
  E(g_mrIML_mutual)$weight < 0,
  yes = "red",
  no = "blue"
)
edge_widths_mrIML_mutual <- abs(E(g_mrIML_mutual)$weight) * 10 # Adjust the scaling factor as needed
matching_indices_mutual <- as.numeric(rownames(mrIML_mat_mutual))

true_mat_mutual <- M_mutual[matching_indices_mutual, matching_indices_mutual]
rownames(true_mat_mutual) <- rownames(mrIML_mat_mutual)
colnames(true_mat_mutual) <- rownames(mrIML_mat_mutual)
g_true_mutual <- graph_from_adjacency_matrix(
  true_mat_mutual,
  mode = 'undirected',
  weighted = T,
  diag = F
)
edge_colors_true_mutual <- ifelse(
  E(g_true_mutual)$weight < 0,
  yes = "red",
  no = "blue"
)
edge_widths_true_mutual <- abs(E(g_true_mutual)$weight) * 5 # Adjust the scaling factor as needed

layout1_mutual <- layout_nicely(g_true_mutual)
```

```{r}
#| results: 'hide'
studyDesign_mutual <- data.frame(sample = factor(1:nrow(Y_mutual)))
rL_mutual <- HmscRandomLevel(units = studyDesign_mutual$sample)
m_hmsc_mutual <- Hmsc(
  Y           = as.matrix(Y_mutual),
  XFormula    = ~1,
  XData       = data.frame(intercept = rep(1, nrow(Y_mutual))),
  distr       = "probit",
  studyDesign = studyDesign_mutual,
  ranLevels   = list("sample" = rL_mutual)
)
m_hmsc_mutual <- sampleMcmc(
  m_hmsc_mutual,
  thin      = 50,
  samples   = 1000,
  transient = 5000,
  nChains   = 4,
  verbose   = 0
)
```

```{r}
mpost_mutual <- convertToCodaObject(m_hmsc_mutual)
psrf_mutual <- gelman.diag(mpost_mutual$Omega[[1]], multivariate = FALSE)$psrf
psrf_mutual
```

```{r}
#| results: 'asis'
if (any(psrf_mutual[, "Point est."] > 1.1)) {
  cat("::: {.callout-warning}\n")
  cat("**HMSC convergence concern:** One or more Omega parameters have Rhat > 1.1,")
  cat(" indicating incomplete chain mixing. Interpret these results cautiously.\n")
  cat(":::\n")
}
```

```{r}
assoc_mutual <- computeAssociations(m_hmsc_mutual)[[1]]
OmegaCor_mutual <- assoc_mutual$mean
OmegaCor_mutual[assoc_mutual$support > 0.05 & assoc_mutual$support < 0.95] <- 0
hmsc_species_names_mutual <- paste0("sp", rownames(mrIML_mat_mutual))
hmsc_mat_mutual <- OmegaCor_mutual[hmsc_species_names_mutual, hmsc_species_names_mutual]
rownames(hmsc_mat_mutual) <- rownames(mrIML_mat_mutual)
colnames(hmsc_mat_mutual) <- rownames(mrIML_mat_mutual)

g_hmsc_mutual <- graph_from_adjacency_matrix(
  hmsc_mat_mutual,
  mode     = "undirected",
  weighted = TRUE,
  diag     = FALSE
)
edge_colors_hmsc_mutual <- ifelse(E(g_hmsc_mutual)$weight < 0, yes = "red", no = "blue")
edge_widths_hmsc_mutual <- abs(E(g_hmsc_mutual)$weight) * 10

par_old <- par(mfrow = c(1, 3))

plot(
  g_mrIML_mutual,
  layout = layout1_mutual,
  edge.color = edge_colors_mrIML_mutual,
  edge.width = edge_widths_mrIML_mutual,
  isolates = TRUE,
  main = "mrIML co-occurrence"
)

plot(
  g_hmsc_mutual,
  layout = layout1_mutual,
  edge.color = edge_colors_hmsc_mutual,
  edge.width = edge_widths_hmsc_mutual,
  isolates = TRUE,
  main = "HMSC co-occurrence"
)

plot(
  g_true_mutual,
  layout = layout1_mutual,
  edge.color = edge_colors_true_mutual,
  isolates = TRUE,
  edge.width = edge_widths_true_mutual,
  main = "Simulated (truth)"
)

par(par_old)
```

```{r}
dist_mat_true_mutual <- 1 - true_mat_mutual
dist_mat_mrIML_mutual <- 1 - mrIML_mat_mutual
mantel_mrIML_mutual <- vegan::mantel(
  as.dist(dist_mat_true_mutual),
  as.dist(dist_mat_mrIML_mutual),
  method = "spearman"
)
mantel_mrIML_mutual

dist_mat_hmsc_mutual <- 1 - hmsc_mat_mutual
mantel_hmsc_mutual <- vegan::mantel(
  as.dist(dist_mat_true_mutual),
  as.dist(dist_mat_hmsc_mutual),
  method = "spearman"
)
mantel_hmsc_mutual
```

## Competitive

```{r}
simulated_compt_network <- generateM_specific_type(
  nn = network_size,
  k_ave = k_average,
  type.network = "random",
  type.interact = "compt",
  interact.str.max = 0.5,
  mix.compt.ratio = 0.5
)
true_compt_network <- simulated_compt_network[[1]]
M_compt <- simulated_compt_network[[2]]
diag(M_compt) <- -0.5 # Reduce the level of self regulation
```

```{r}
data_compt <- generateDataSet(
  900,
  M_compt,
  count = network_size * 10000,
  mode = 4
) %>%
  t() %>%
  as.data.frame() %>%
  mutate(
    across(everything(), ~ ifelse(. < 1, yes = 0, no = 1))
  )
```

```{r}
#| results: 'hide'
# Prepare data
Y_compt <- filterRareCommon(data_compt, lower = 0.01, higher = 0.99)
X1_compt <- Y_compt
# Fit mrIML obj
yhats_rf_sim_compt <- mrIMLpredicts(
  Y = Y_compt,
  X = NULL,
  X1 = X1_compt,
  Model = model_rf,
  prop = 0.7,
  k = 5,
  racing = FALSE
)
# Bootstrapping
bs_sim_compt <- mrBootstrap(yhats_rf_sim_compt)
# Extract network
assoc_net_compt <- mrCoOccurNet(bs_sim_compt)
```

```{r}
mrIML_mat_compt <- assoc_net_compt %>%
  filter(mean_strength > 0.1) %>%
  mutate(
    mean_strength_dir = ifelse(
      direction == "negative",
      yes = -mean_strength,
      no = mean_strength
    ),
    across(contains("taxa"), ~ sub("^sp", "", .))
  ) %>%
  graph_from_data_frame(
    directed = FALSE,
    vertices = sub("^sp", "", names(Y_compt))
  ) %>%
  get.adjacency(attr = "mean_strength_dir", sparse = FALSE)

#plots
g_mrIML_compt <- graph_from_adjacency_matrix(
  mrIML_mat_compt,
  mode = 'undirected',
  weighted = T,
  diag = F
)

edge_colors_mrIML_compt <- ifelse(
  E(g_mrIML_compt)$weight < 0,
  yes = "red",
  no = "blue"
)
edge_widths_mrIML_compt <- abs(E(g_mrIML_compt)$weight) * 10 # Adjust the scaling factor as needed
matching_indices_compt <- as.numeric(rownames(mrIML_mat_compt))

true_mat_compt <- M_compt[matching_indices_compt, matching_indices_compt]
rownames(true_mat_compt) <- rownames(mrIML_mat_compt)
colnames(true_mat_compt) <- rownames(mrIML_mat_compt)
g_true_compt <- graph_from_adjacency_matrix(
  true_mat_compt,
  mode = 'undirected',
  weighted = T,
  diag = F
)
edge_colors_true_compt <- ifelse(
  E(g_true_compt)$weight < 0,
  yes = "red",
  no = "blue"
)
edge_widths_true_compt <- abs(E(g_true_compt)$weight) * 5 # Adjust the scaling factor as needed

layout1_compt <- layout_nicely(g_true_compt)
```

```{r}
#| results: 'hide'
studyDesign_compt <- data.frame(sample = factor(1:nrow(Y_compt)))
rL_compt <- HmscRandomLevel(units = studyDesign_compt$sample)
m_hmsc_compt <- Hmsc(
  Y           = as.matrix(Y_compt),
  XFormula    = ~1,
  XData       = data.frame(intercept = rep(1, nrow(Y_compt))),
  distr       = "probit",
  studyDesign = studyDesign_compt,
  ranLevels   = list("sample" = rL_compt)
)
m_hmsc_compt <- sampleMcmc(
  m_hmsc_compt,
  thin      = 50,
  samples   = 1000,
  transient = 5000,
  nChains   = 4,
  verbose   = 0
)
```

```{r}
mpost_compt <- convertToCodaObject(m_hmsc_compt)
psrf_compt <- gelman.diag(mpost_compt$Omega[[1]], multivariate = FALSE)$psrf
psrf_compt
```

```{r}
#| results: 'asis'
if (any(psrf_compt[, "Point est."] > 1.1)) {
  cat("::: {.callout-warning}\n")
  cat("**HMSC convergence concern:** One or more Omega parameters have Rhat > 1.1,")
  cat(" indicating incomplete chain mixing. Interpret these results cautiously.\n")
  cat(":::\n")
}
```

```{r}
assoc_compt <- computeAssociations(m_hmsc_compt)[[1]]
OmegaCor_compt <- assoc_compt$mean
OmegaCor_compt[assoc_compt$support > 0.05 & assoc_compt$support < 0.95] <- 0
hmsc_species_names_compt <- paste0("sp", rownames(mrIML_mat_compt))
hmsc_mat_compt <- OmegaCor_compt[hmsc_species_names_compt, hmsc_species_names_compt]
rownames(hmsc_mat_compt) <- rownames(mrIML_mat_compt)
colnames(hmsc_mat_compt) <- rownames(mrIML_mat_compt)

g_hmsc_compt <- graph_from_adjacency_matrix(
  hmsc_mat_compt,
  mode     = "undirected",
  weighted = TRUE,
  diag     = FALSE
)
edge_colors_hmsc_compt <- ifelse(E(g_hmsc_compt)$weight < 0, yes = "red", no = "blue")
edge_widths_hmsc_compt <- abs(E(g_hmsc_compt)$weight) * 10

par_old <- par(mfrow = c(1, 3))

plot(
  g_mrIML_compt,
  layout = layout1_compt,
  edge.color = edge_colors_mrIML_compt,
  edge.width = edge_widths_mrIML_compt,
  isolates = TRUE,
  main = "mrIML co-occurrence"
)

plot(
  g_hmsc_compt,
  layout = layout1_compt,
  edge.color = edge_colors_hmsc_compt,
  edge.width = edge_widths_hmsc_compt,
  isolates = TRUE,
  main = "HMSC co-occurrence"
)

plot(
  g_true_compt,
  layout = layout1_compt,
  edge.color = edge_colors_true_compt,
  isolates = TRUE,
  edge.width = edge_widths_true_compt,
  main = "Simulated (truth)"
)

par(par_old)
```

```{r}
dist_mat_true_compt <- 1 - true_mat_compt
dist_mat_mrIML_compt <- 1 - mrIML_mat_compt
mantel_mrIML_compt <- vegan::mantel(
  as.dist(dist_mat_true_compt),
  as.dist(dist_mat_mrIML_compt),
  method = "spearman"
)
mantel_mrIML_compt

dist_mat_hmsc_compt <- 1 - hmsc_mat_compt
mantel_hmsc_compt <- vegan::mantel(
  as.dist(dist_mat_true_compt),
  as.dist(dist_mat_hmsc_compt),
  method = "spearman"
)
mantel_hmsc_compt
```

## Mixed

```{r}
simulated_mix_network <- generateM_specific_type(
  nn = network_size,
  k_ave = k_average,
  type.network = "random",
  type.interact = "mix",
  interact.str.max = 0.4,
  mix.compt.ratio = 0.5
)
true_mix_network <- simulated_mix_network[[1]]
M_mix <- simulated_mix_network[[2]]
```

```{r}
data_mix <- generateDataSet(
  900,
  M_mix,
  count = network_size * 10000,
  mode = 4
) %>%
  t() %>%
  as.data.frame() %>%
  mutate(
    across(everything(), ~ ifelse(. < 1, yes = 0, no = 1))
  )
```

```{r}
#| results: 'hide'
# Prepare data
Y_mix <- filterRareCommon(data_mix, lower = 0.01, higher = 0.99)
X1_mix <- Y_mix
# Fit mrIML obj
yhats_rf_sim_mix <- mrIMLpredicts(
  Y = Y_mix,
  X = NULL,
  X1 = X1_mix,
  Model = model_rf,
  prop = 0.7,
  k = 5,
  racing = FALSE
)
# Bootstrapping
bs_sim_mix <- mrBootstrap(yhats_rf_sim_mix)
# Extract network
assoc_net_mix <- mrCoOccurNet(bs_sim_mix)
```

```{r}
mrIML_mat_mix <- assoc_net_mix %>%
  filter(mean_strength > 0.1) %>%
  mutate(
    mean_strength_dir = ifelse(
      direction == "negative",
      yes = -mean_strength,
      no = mean_strength
    ),
    across(contains("taxa"), ~ sub("^sp", "", .))
  ) %>%
  graph_from_data_frame(
    directed = FALSE,
    vertices = sub("^sp", "", names(Y_mix))
  ) %>%
  get.adjacency(attr = "mean_strength_dir", sparse = FALSE)

#plots
g_mrIML_mix <- graph_from_adjacency_matrix(
  mrIML_mat_mix,
  mode = 'undirected',
  weighted = T,
  diag = F
)

edge_colors_mrIML_mix <- ifelse(
  E(g_mrIML_mix)$weight < 0,
  yes = "red",
  no = "blue"
)
edge_widths_mrIML_mix <- abs(E(g_mrIML_mix)$weight) * 10 # Adjust the scaling factor as needed
matching_indices_mix <- as.numeric(rownames(mrIML_mat_mix))

true_mat_mix <- M_mix[matching_indices_mix, matching_indices_mix]
rownames(true_mat_mix) <- rownames(mrIML_mat_mix)
colnames(true_mat_mix) <- rownames(mrIML_mat_mix)
g_true_mix <- graph_from_adjacency_matrix(
  true_mat_mix,
  mode = 'undirected',
  weighted = T,
  diag = F
)
edge_colors_true_mix <- ifelse(
  E(g_true_mix)$weight < 0,
  yes = "red",
  no = "blue"
)
edge_widths_true_mix <- abs(E(g_true_mix)$weight) * 5 # Adjust the scaling factor as needed

layout1_mix <- layout_nicely(g_true_mix)
```

```{r}
#| results: 'hide'
studyDesign_mix <- data.frame(sample = factor(1:nrow(Y_mix)))
rL_mix <- HmscRandomLevel(units = studyDesign_mix$sample)
m_hmsc_mix <- Hmsc(
  Y           = as.matrix(Y_mix),
  XFormula    = ~1,
  XData       = data.frame(intercept = rep(1, nrow(Y_mix))),
  distr       = "probit",
  studyDesign = studyDesign_mix,
  ranLevels   = list("sample" = rL_mix)
)
m_hmsc_mix <- sampleMcmc(
  m_hmsc_mix,
  thin      = 50,
  samples   = 1000,
  transient = 5000,
  nChains   = 4,
  verbose   = 0
)
```

```{r}
mpost_mix <- convertToCodaObject(m_hmsc_mix)
psrf_mix <- gelman.diag(mpost_mix$Omega[[1]], multivariate = FALSE)$psrf
psrf_mix
```

```{r}
#| results: 'asis'
if (any(psrf_mix[, "Point est."] > 1.1)) {
  cat("::: {.callout-warning}\n")
  cat("**HMSC convergence concern:** One or more Omega parameters have Rhat > 1.1,")
  cat(" indicating incomplete chain mixing. Interpret these results cautiously.\n")
  cat(":::\n")
}
```

```{r}
assoc_mix <- computeAssociations(m_hmsc_mix)[[1]]
OmegaCor_mix <- assoc_mix$mean
OmegaCor_mix[assoc_mix$support > 0.05 & assoc_mix$support < 0.95] <- 0
hmsc_species_names_mix <- paste0("sp", rownames(mrIML_mat_mix))
hmsc_mat_mix <- OmegaCor_mix[hmsc_species_names_mix, hmsc_species_names_mix]
rownames(hmsc_mat_mix) <- rownames(mrIML_mat_mix)
colnames(hmsc_mat_mix) <- rownames(mrIML_mat_mix)

g_hmsc_mix <- graph_from_adjacency_matrix(
  hmsc_mat_mix,
  mode     = "undirected",
  weighted = TRUE,
  diag     = FALSE
)
edge_colors_hmsc_mix <- ifelse(E(g_hmsc_mix)$weight < 0, yes = "red", no = "blue")
edge_widths_hmsc_mix <- abs(E(g_hmsc_mix)$weight) * 10

par_old <- par(mfrow = c(1, 3))

plot(
  g_mrIML_mix,
  layout = layout1_mix,
  edge.color = edge_colors_mrIML_mix,
  edge.width = edge_widths_mrIML_mix,
  isolates = TRUE,
  main = "mrIML co-occurrence"
)

plot(
  g_hmsc_mix,
  layout = layout1_mix,
  edge.color = edge_colors_hmsc_mix,
  edge.width = edge_widths_hmsc_mix,
  isolates = TRUE,
  main = "HMSC co-occurrence"
)

plot(
  g_true_mix,
  layout = layout1_mix,
  edge.color = edge_colors_true_mix,
  isolates = TRUE,
  edge.width = edge_widths_true_mix,
  main = "Simulated (truth)"
)

par(par_old)
```

```{r}
dist_mat_true_mix <- 1 - true_mat_mix
dist_mat_mrIML_mix <- 1 - mrIML_mat_mix
mantel_mrIML_mix <- vegan::mantel(
  as.dist(dist_mat_true_mix),
  as.dist(dist_mat_mrIML_mix),
  method = "spearman"
)
mantel_mrIML_mix

dist_mat_hmsc_mix <- 1 - hmsc_mat_mix
mantel_hmsc_mix <- vegan::mantel(
  as.dist(dist_mat_true_mix),
  as.dist(dist_mat_hmsc_mix),
  method = "spearman"
)
mantel_hmsc_mix
```

## Summary

Comparison of Mantel r (Pearson) and p-values for mrIML and HMSC against the true interaction network across all simulation scenarios.

```{r}
comparison_table <- tibble(
  Scenario = c("Random", "Mutualistic", "Competitive", "Mixed"),
  mrIML_r  = round(c(
    mantel_mrIML$statistic,
    mantel_mrIML_mutual$statistic,
    mantel_mrIML_compt$statistic,
    mantel_mrIML_mix$statistic
  ), 3),
  mrIML_p  = c(
    mantel_mrIML$signif,
    mantel_mrIML_mutual$signif,
    mantel_mrIML_compt$signif,
    mantel_mrIML_mix$signif
  ),
  HMSC_r   = round(c(
    mantel_hmsc$statistic,
    mantel_hmsc_mutual$statistic,
    mantel_hmsc_compt$statistic,
    mantel_hmsc_mix$statistic
  ), 3),
  HMSC_p   = c(
    mantel_hmsc$signif,
    mantel_hmsc_mutual$signif,
    mantel_hmsc_compt$signif,
    mantel_hmsc_mix$signif
  )
)

DT::datatable(
  comparison_table,
  colnames = c("Scenario", "mrIML r", "mrIML p", "HMSC r", "HMSC p"),
  options  = list(pageLength = 10, dom = "t")
)
```

# Asymetric simulations

...